\documentclass[a4paper,11pt,english]{article}
\usepackage{.style}

%%%%%%% Title %%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\textbf{Topology II - Homology vanishing above the dimension}}
\author{Tor Gjone}

%%%%%%% Definitions %%%%%%%%%%%%%%%%%%%%%
\input{defs}

\DeclareMathOperator{\Gr}{Gr}

%%%%%%%% Content %%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\mmaketitle

% NOT Complete
% Missing proofs 

All of the manifolds that we explicitly discussed admit CW-structures:

\begin{itemize}
\item $S^n$, $\RP^n$, $\CP^n$ and $\HP^n$,
\item $V_{k,n}$, $\Gr_{k,n}$: CW-structures exists.
\end{itemize}

In all these cases: manifold dimension = CW-dimension

This is no coincidence: Suppose that a compact manifold $M$ admits a
CW-structure; let $x\in M$ be an interior point of a cell of top dimension 
(in the CW-structure). Then the open cell is is an open neighbourhood
homeomorphic to $\R^n$ with $n = $ CW-dimension of $M$. Since the manifold
dimension is intrinsic, $n =$ manifold dimension.

\begin{corol}
Let $M$ be a compact $n$-manifold that admits a CW-structure. Then
$H_i(M; A) \cong H_i^{\text{cell}}(M;A) = 0$ for $i > n$.
\end{corol}

Warning: compact manifolds do not in general admit CW-structures. But every
smooth compact manifold admits a triangulation, and hence a CW-structure.

Notation: For $M$ an $n$-manifold, $A$ an abelian group and $U \subset M$, let $H_i(M|U; A) =
H_i(M, M \setminus U; A)$ denote the local $i$'th homology at $U$ with values in $A$.

\begin{thrm}
Let $M$ be an $n$-manifold, $A$ and abelian group and $K$ a compact subset of
$M$. Then
\begin{enumerate}
\item[(i)] $H_i(M|K; A) = 0$ for $i > n$.
\item[(ii)] A class in $H_n(M|K; A)$ is zero if and only if its
restriction to $H_n(M|x; A)$ is zero for all $x \in K$.
\end{enumerate}
\end{thrm}

\begin{note}
$M$ needs not be compact. But if $M$ is compact, then $K = M$ is allowed and
then both statements refer to absolute homology of $M$.
\end{note}


\begin{proof}
(The proof follows the proof of Lemma A.7 in Appendix A of Milnor-Stasheff's
book "Characteristic Classes")

The proof is done in 6 steps.

\begin{enumerate}
\item[(Step 1)] Consider $M = \R^n$, $K$ is a compact convex non-empty subset.
For every $x \in K$, $K$ can be linearly contracted onto $x$. Let $R > 0$ be
large enough so that $K \subseteq B^{n-1}_R(x) := \{ y \in \R^n \q: |x-y| \le R
\}$. Then the inclusion
\[ S^{-1}_{2R}(x) := \{ y \in \R^n \q: |x-y| = 2R \} \subseteq M\setminus K
\subseteq M \setminus \{ x \}, \]
defines homotopy equivalences. So the induced map $H_i(M|K) \to
H_i(M|x)$ is an isomorphism.

In particular, $H_i(M|K) \cong H_i(M|x) \cong 0$ for $i > n$.

\item[(Step 2)]
Let $M$ be any $n$-manifold, $K = K_1 \cup K_2$ for $K_!, K_2$ compact, suppose
the statements are true for $K_1, K_2$ and $K_1 \cap K_2$. Then the statements
also hold for $K$. We have a long exact Mayer-Vietro's sequence for the local
homology groups. 

\underline{Construction:}
\[ M \setminus (K_1 \cap K_2) = (M\setminus K_1) \cup (M \setminus K_2)
\qq{\text{and}} (M\setminus K_1) \cap (M \setminus K_2) = M \setminus (K_1 \cup
K_2) = M \setminus K. \]
The theorem of small subjections show that the map 
\[ \frac{C_*(M\setminus K_1) \oplus C_*(M\setminus K_2)}{C_*(M\setminus K)}
\incto[\cong] C_*(M \setminus (K_1 \cap K_2)) \]
is an isomorphism of all homology groups.

So the chain map
\[  D := \frac{C_*(M)}{ \left( \frac{C_*(M\setminus K_1) \oplus C_*(M\setminus
K_2)}{C_*(M\setminus K)} \right) } \congto \frac{C_*(M)}{C_*(M\setminus
(K_1 \cap K_2))} \]
is also a quasi-isomorphism.

The so
\end{enumerate}
\end{proof}

\end{document}
