\documentclass[a4paper,11pt,english]{article}
\usepackage{.style}

%%%%%%% Title %%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{\textbf{Topology II - The fundamental class}}
\author{Tor Gjone}

%%%%%%% Definitions %%%%%%%%%%%%%%%%%%%%%
\input{defs}

\newcommand{\bM}{\bar{M}}
\DeclareMathOperator{\supp}{supp}

%%%%%%%% Content %%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\mmaketitle

% NOT Complete
% Missing proofs 

Homology of some connected compact manifolds: 

\begin{align*}
H_n(S^n;\Z) &\cong H_{2n}(\CP^n; \Z) \cong H_{4n}(\HP^n;\Z) \cong \Z \\
H_n(\RP^n; \Z) &\cong \begin{cases} \Z &\tif n \text{ is odd} \\ 0 &\tif n \text{ is even} \end{cases}
\end{align*}

We will show locally that for a connected, compact, non-empty $n$-manifold
$M$, 
\[ H_n(M; \Z) \cong \begin{cases} \Z &\tif M \text{ is orientable} \\ 0 &\tif n
\text{ is not orientable} \end{cases} \]

\begin{thrm}
Let $M$ be an orientable $n$-manifold and let $K$ ve a compat subset of
$M$. Then there is a unique class $\mu_K \in H_n(M|K)$ such that
$r^K_x(\mu_K) = \mu_x$ gives local orientation in $H_n(M|x)$ for all $x\in K$. 
\end{thrm}

\subsection{Important special calse:} 
If $M$ is it self compact, we can take $K = M$; there is then a unique class
$\mu_M = [M] \in H_n(M;\Z)$, the \wddef{fundamental class} such that
$r^M_x[M] = \mu_x$ in $H_n(M|x)$ for all $x\in M$.

We will show later that if $M$ is moreover connected and non-empty then
$H_n(M; \Z) \cong \Z$ with generator $[M]$.


\begin{remark}
Let $M$ be orientable, compact, connected. Let $\bar{M}$ be $M$ with the
opposite orientation. Then $[\bM] = -[M]$, because both have the same image in
$H_n(M|x)$ for all $x \in M$.
\end{remark}

\begin{remark}
Compactness is necessary: Let $M$ be an $n$-manifold and $U \in H_n(M; A)$
(for $A$ some abelian group) such that $r^M_x(u) \ne 0$ in $H_n(M|x;A)$ for all
$x \in M$. Then $M$ is compact. Indeed, write $u = [x]$ for some $n$-cycle $x
\in C_n(M;A)$. Write $x \in \sum_{\text{finite}} a_i (f_i : \nabla^n \to M)$,
and set $L = \supp(x) = \cup f_i(\nabla^n)$, some compact subset of $M$. 

If $M$ is not compact, there is a point $x \in M \setminus L$. So $x \in
C_n(M \setminus \{ x \}; A)$ is a cycle in $M \setminus \{ x \}$. So $u = [x]$
has zero image in $H_n(M|x;A)$ This contradicts the assumption on $u$.
\end{remark}

\begin{proof}
(following the proof of Theorem A.8 in Appendix A of Milnor-Stasheff's
"Characteristic Classes")

Yet to be typed up.
\end{proof}

\begin{corol}
Let $M$ be an orientable, compact, connected $n$-manifold. Then for all $x\in
M$, the map 
\[ r^M_x : H_n(M;\Z) \to H_n(M|x) \]
is an isomorphism. So if $M$ is non-empty, $H_n(M;\Z)$ is free of rank 1, and
the fundamental class of the two orientations are the two generators.
\end{corol}

\begin{proof}
To Be done
\end{proof}

\begin{corol}
Let $M$ be a compact, connected $n$-manifold that is not orientable. Then
$H_n(M;\Z) = 0$.
\end{corol}

\end{document}
